---
title: "EVT on Synthetic Data (Max Domain of Attarction)"
author: "Rishikesh Yadav"
date: "2024-08-07"
output: html_document
---

## Introduction

This document demonstrates the simulation of data from different distributions, extraction of yearly maxima, and fitting of Generalized Extreme Value (GEV) and Generalized Pareto Distribution (GPD) using the `extRemes` package.

## Setup

```{r setup, include=FALSE}
library(extRemes)
library(evd)
library(VGAM)  # Load VGAM package for Pareto distribution
set.seed(123)  # Set seed for reproducibility
```

## Simulate Data

We simulate 100 years of data from various distributions, excluding the uniform distribution.

```{r simulate-data}
# Number of years and number of data points per year
num_years <- 100
data_per_year <- 365

# Simulate data for each distribution
cauchy_data <- matrix(rcauchy(num_years * data_per_year), nrow = num_years)
gaussian_data <- matrix(rnorm(num_years * data_per_year), nrow = num_years)
exponential_data <- matrix(rexp(num_years * data_per_year), nrow = num_years)
pareto_data <- matrix(rpareto(num_years * data_per_year, shape = 1), nrow = num_years)  # using VGAM package

# Extract yearly maxima
max_cauchy <- apply(cauchy_data, 1, max)
max_gaussian <- apply(gaussian_data, 1, max)
max_exponential <- apply(exponential_data, 1, max)
max_pareto <- apply(pareto_data, 1, max)
```

## Fit GEV Distribution

Fit GEV distribution to the yearly maxima data.

```{r fit-gev}
# Fit GEV distribution
fit_gev_cauchy <- fevd(max_cauchy, type = "GEV")
fit_gev_gaussian <- fevd(max_gaussian, type = "GEV")
fit_gev_exponential <- fevd(max_exponential, type = "GEV")
fit_gev_pareto <- fevd(max_pareto, type = "GEV")

sumry_gev_cauchy<- summary(fit_gev_cauchy)
sumry_gev_gaussian<- summary(fit_gev_gaussian)
sumry_gev_exponential<- summary(fit_gev_exponential)
sumry_gev_pareto<- summary(fit_gev_pareto)

```